Demystifying the Computer Science PhD Admission in the US\\{\large A Guide for International Students

### 2023-11-09 demystify.tex
updates

\title{Demystifying the Computer Science PhD Admission\\in US Universities\\{\large A Guide for International Students}}

@@

\makeatletter
\def\maketitle{%
  \null
  \thispagestyle{empty}%
  \vfill
  \begin{center}\leavevmode
    \normalfont
    {\LARGE\raggedright \@title\par}%
    \vfill%
    %\hrulefill\par
    {\Large \@author\par}%
    \vfill%
    {\large\raggedleft \@date\par}%
  \end{center}%
  \vfill
  \null
  \cleardoublepage
  }
\makeatother
%\author{ThanhVu H. Nguyen}
%\title{TITLE OF BOOK}
%\date{}

@@

\chapter{Summary}\label{sec:summary}
Here we summarize the main points of the document. This allows you to quickly skim through the document and decide which parts you want to read in detail.
\begin{enumerate}
\item Should you apply to a CS PhD in the US?
\begin{itemize}
    \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\S\ref{sec:should}).
\end{itemize}
\item How is your application evaluated?
\begin{itemize}
    \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically 3 faculty (\S\ref{sec:evalapps}).
    \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so contact and introduce yourself (\S\ref{sec:contact}).
\end{itemize}
\item Application Materials
\begin{itemize}
    \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, personal statement, and recommendation letters.
    \item LORs are important, but only if they are personalized and talk about your research ability (\S\ref{sec:lor}).
    \item GRE \emph{is not} required by most schools (\S\ref{sec:grades}).
    \item Grades are important, but only if your school is well-known (\S\ref{sec:grades}).
    \item Personal statement is very important. Write it in such a way that makes you \emph{stand out} (\S\ref{sec:research-statement} and \S\ref{sec:improve-your-chance})
    \item Getting an interview is typically a \emph{good sign} (no one interviews weak candidates).
\end{itemize}
\item What to do after getting admitted?
\begin{itemize}
    \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\S\ref{sec:accepted}).
\end{itemize}
\item Funding
\begin{itemize}
    \item TA and RA are two main sources of funding.  TA (teaching assistantship) is provided by the department to help profs. with classes (e.g., grading). RA (research assistantship) is provided by profs. to help them with their research (\S\ref{sec:funding}).
\end{itemize}
\item Choosing School and Professors
\begin{itemize}
    \item Many schools do not offer PhD studies in CS and many CS professors \emph{cannot} be adviser, i.e., cannot graduate PhD students  (\S\ref{sec:schoolsandprofs}).
    \item Contacting a prof. for research opportunities is recommended, but do it \emph{properly} (\S\ref{sec:contact}).
\end{itemize}
\item Miscs and FAQS
\begin{itemize}
    \item Increasing your admission chance by being unique and standing out (\S\ref{sec:improve-your-chance}).
    \item You can successfully apply to CS PhD even if you have non-STEM background (\S\ref{sec:non-stem}).
    \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\S\ref{sec:non-us-differences}).
    \item Your PhD costs about \$400K in total, but you do not pay for it (\S\ref{sec:ra-cost}).
    \item Despite some miserable stories on social media, many PhD students have good mentors, supportive labmates, healthy working environment ... and are happy (\S\ref{sec:happy}).
\end{itemize}
\end{enumerate}

@@

Second, I believe that applying to a good US university \emph{should not} be any
harder than at schools in other countries. If you believe you have a chance in other countries, e.g., South Korea, Singapore, Germany, UK, Japan and Australia, then you will surely have a chance in the US as well\footnote{\S\ref{sec:non-us-differences} compares CS PhD study in the US to other countries.}.

\begin{commentbox}[Vu:]
One of the reasons I create this document is that my colleagues at GMU are interested in recruiting Vietnamese students and are surprised when seeing very few applications in Vietnam (e.g., each year our CS program receives more than 350 PhD applications, most of which are international but only 3--4 are from Vietnam). In general the number of CS PhD applications from Vietnam to US universities is low and more would be welcomed. 
\end{commentbox}

@@

\chapter{Application}\label{sec:application}

@@

\section{Research Ability}

@@

    In other fields, e.g. Software Engineering, Vietnamese students face challenges in reaching US professors. Do you have any tips for Vietnamese students who want to connect with US professors and work as research assistants? 
    \tcblower
    \textbf{Vu}: \S\ref{sec:contact} shows how to contact a professor for research opportunities. Many probably will say no as they do not have the bandwidth to take on random students, but some may say yes if they see potential fits.  
\end{commentbox}

@@

Plagiarism and cheating (e.g., exams, coding) is a BIG no-no in the US.  If you're caught cheating, you will face very heavy consequence and likely be expelled from the university (e.g., after the second time at GMU).   This is quite different from many international countries where cheating is common and often tolerated.  Faculty is extremely good at detecting cheating (we have been dealing with these situations so many times over so many years), and \emph{will} report cheating cases.  In short, whatever you do, don't cheat---not worth it.

Here is the typically steps: (i) a faculty suspecting a cheating case will report it to the Office of Academic Integrity (OAI) at the university---the report often has supporting evidence and suggested penalty (e.g., receive a failing grade);  (ii)  OAI will take over and investigate the case; and (iii) OAI will make the final decision.  It is important to note that after receiving the report from your prof., OAI \emph{completely} takes over and makes decision.  This means begging your professor will not help because they simply are no longer involved in the case.

@@

\chapter{History and Acknowledgement}

>>>>

\title{Demystifying the Computer Science PhD Admission in the US\\{\large A Guide for International Students}}

@@

\makeatletter
\def\maketitle{%
  \null
  \thispagestyle{empty}%
  \vfill
  \begin{center}\leavevmode
    \normalfont
    {\LARGE\raggedright \textbf{\@title}\par}%
    \vfill%
    %\hrulefill\par
    {\Large \@author\par}%
    \vfill%
    {\large\raggedleft \@date\par}%
  \end{center}%
  \vfill
  \null
  \cleardoublepage
  }
\makeatother
%\author{ThanhVu H. Nguyen}
%\title{TITLE OF BOOK}
%\date{}

@@

\chapter{Summary}\label{sec:summary}
Here we summarize the main points of the document. This allows you to quickly skim through the document and decide which parts you want to read in detail.
\begin{enumerate}
\item Should you apply to a CS PhD in the US?
\begin{itemize}
    \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\S\ref{sec:should}).
\end{itemize}
\item How is your application evaluated?
\begin{itemize}
    \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically 3 faculty (\S\ref{sec:evalapps}).
    \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so contact and introduce yourself (\S\ref{sec:contact}).
\end{itemize}
\item Application Materials
\begin{itemize}
    \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, personal statement, and recommendation letters.
    \item LORs are very important, but only if they are personalized and talk about your research ability (\S\ref{sec:lor}).
    \item Personal statement is very important. Write it in such a way that makes you \emph{stand out} (\S\ref{sec:research-statement} and \S\ref{sec:improve-your-chance})
    \item GRE \emph{is not} required (\S\ref{sec:grades}). Spend your time on something else.
    \item Grades are important, but depend on the reputation of your school (\S\ref{sec:grades}).
    \item Getting an interview is typically a \emph{good sign} (no one interviews weak candidates).
\end{itemize}
\item What to do after getting admitted?
\begin{itemize}
    \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\S\ref{sec:accepted}).
\end{itemize}
\item Funding
\begin{itemize}
    \item TA and RA are two main sources of funding.  TA (teaching assistantship) is provided by the department to help profs. with classes (e.g., grading). RA (research assistantship) is provided by profs. to help them with their research (\S\ref{sec:funding}).
\end{itemize}
\item Choosing School and Professors
\begin{itemize}
    \item Many schools do not offer PhD studies in CS and many CS professors \emph{cannot} be adviser, i.e., cannot graduate PhD students  (\S\ref{sec:schoolsandprofs}).
    \item Contacting a prof. for research opportunities is recommended, but do it \emph{properly} (\S\ref{sec:contact}).
\end{itemize}
\item Miscs and FAQS
\begin{itemize}
    \item Increasing your admission chance by being unique and standing out (\S\ref{sec:improve-your-chance}).
    \item You can successfully apply to CS PhD even if you have non-STEM background (\S\ref{sec:non-stem}).
    \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\S\ref{sec:non-us-differences}).
    \item Your PhD costs about \$400K in total, but you do not pay for it (\S\ref{sec:ra-cost}).
    \item Despite some miserable stories on social media, many PhD students have good mentors, supportive labmates, healthy working environment ... and are happy (\S\ref{sec:happy}).
\end{itemize}
\end{enumerate}

@@

Second, I believe that applying to a good US university \emph{should not} be any
harder than at schools in other countries. If you believe you have a chance in other countries, e.g., South Korea, Singapore, Germany, UK, Japan and Australia, then you will surely have a chance in the US as well. \S\ref{sec:non-us-differences} compares CS PhD study in the US to other countries.

\begin{commentbox}[Vu:]
One of the reasons I create this document is that my colleagues at GMU are interested in recruiting Vietnamese students and are surprised when seeing very few applications in Vietnam (see \S\ref{sec:ack}). Each year our CS program receives more than 350 PhD applications, most of which are international but only 3--4 are from Vietnam. In general the number of CS PhD applications from Vietnam to US universities is very low and more would be welcome. 
\end{commentbox}

@@

\chapter{Your Application}\label{sec:application}

@@

\section{Research Ability}\label{sec:research-ability}

@@

    In other fields, e.g. Software Engineering, Vietnamese students face challenges in reaching US professors. Do you have any tips for Vietnamese students who want to connect with US professors and work as research assistants? 
    \tcblower
    \textbf{Vu}: \S\ref{sec:contact} shows how to contact a professor for research opportunities. Many probably will say no as they do not have the bandwidth to take on random students, but some may say yes if they see potential fit.  
\end{commentbox}

@@

Plagiarism and cheating (e.g., exams, assignments) is a BIG no-no in the US.  If you're caught cheating, you will face very heavy consequence and likely be expelled from the university (e.g., after the second time at GMU).   This is quite different from many international countries where cheating is common and often tolerated.  Faculty is extremely good at detecting cheating (we have been dealing with these situations so many times over so many years), and \emph{will} report cheating cases.  In short, whatever you do, don't cheat---not worth it.

Here is the typically steps: (i) a faculty suspecting a cheating case will report it to the Office of Academic Integrity (OAI) at the university---the report often has supporting evidence and suggested penalty (e.g., a failing grade);  (ii)  OAI will take over and investigate the case; and (iii) OAI will make the final decision.  It is important to note that after receiving the report from your prof., OAI \emph{completely} takes over and makes decision.  This means begging your professor will not help because they simply are no longer involved in the case.

@@

\chapter{History and Acknowledgement}\label{sec:ack}
